The Second Data Release of the INT Photometric H$\alpha$ Survey of the Northern Galactic Plane (IPHAS DR2)

### 2014-07-28 paper/referee/response-revision1.txt

MN-14-1011-MJ: Authors' response to the reviewer's comments
===========================================================

We are grateful to the referee for the insightful feedback, which has helped us
to improve the presentation of the data products.  In his/her report, the 
referee identified areas where additional information would (i) strengthen the 
role of the paper as a user's guide to the survey, and (ii) would lessen the 
burden on the reader to track down information from previous works.  
We appreciate the merits of the referee's suggestions in this respect, and have 
acted upon them by adding a total of N new pages, 1 new table, and 1 new figure 
to the manuscript.  These changes have been marked in bold in the manuscript.  
We have also added N new sections of auxillary material to our web site as a 
direct consequence.

In what follows we will start by explaining our response to each of the issues 
which the referee identified as high-level remarks (9 items), followed by the
referee's additional comments which affect specific sections (8 items).

@@

(There does not appear to be an item number 8 in the reviewer's comments.)

[Referee]
> 9. The authors point out in the introduction that IPHAS will complement a number of
> new infrared surveys of the Galactic Plane.  The paper would benefit from at least
> one cross-comparison with the catalogue of one of those surveys, such as GLIMPSE
> or MIPSGAL, as an illustration of how the combination of surveys will enable
> new science.  At present, the only tie-in to the infrared surveys is the
> visual comparison between IPHAS and GLIMPSE images of one field.

[Authors' response]
In response to the referee's suggestion to include a cross-comparison between
IPHAS and an infrared survey, we have added a brief discussion on the overlap
between the GLIMPSE and IPHAS catalogues at the end of Section 7.2.  The 
discussion warns the reader that an optical survey will naturally be 
extinction-limited.

Section-specific remarks
------------------------

[Referee]
> 1. In Section 3.1, paragraph 1, sentence 1:
> "archival" should be "archiving"

[Authors' response]
This has been corrected.

[Referee]
> 2. In the second paragraph of section 3.1, it is stated that a deblending routine
> is used to remove contaminating sources.  Additional information about this routine
> should be provided.  Are contaminants removed from the sky background measurement,
> the source measurement aperture, both?

[Referee]
> 3. At the beginning of the Data Reduction section (section 3), it would be helpful
> to provide a summary of how data processing for this release differed from that
> used for the Initial Data Release.

[Referee]
> 4.a. It is stated in section 3.2 that the exposures were discarded unless they
> had 5-sigma limiting magnitudes of r>20 or i>19 and H-apha>19 mag. How were the
> limiting magnitudes for an exposure estimated?

[Referee]
> 4.b. Section 3.2 also specifies that an ellipticity limit of eps>0.3 was used.
> What is the justification of this limit?  For example, what is the potential impact on
> photometry and sensitivity at eps=0.25?

[Referee]
> 5.a. Were the atmospheric extinction coefficients of 0.09 (r) and 0.05 (i) simply
> adopted, or was there any attempt to measure them?  Is there an estimate of
> how much the extinction may have varied with time, either during a night, or
> seasonally?

[Referee]
> 5.b. The PERCORR factor in equation 1 is used to correct for "sporadic gain variation."
> What was the nature of the variations when non-zero?  How was it measured?
> What is the size of the correction when it was applied?

[Referee]
> 5.c. The first paragraph of section 4.2 on Global Calibration states that calibration
> errors  can be up to 0.1 magnitude in reference to the colour-colour diagram shown in
> Figure 4a. There are outlier populations in that figure that are up to 0.6 mags
> from the main locus of points.  These seem to be accurately corrected by
> the recalibration, but the text seems to understate the severity of the
> systematics.

[Referee]
> 5.d.  It is stated in section 4.2, paragraph 2 that 2MASS observed six calibration
> fields each hour.  That is not correct.  2MASS observed two calibration fields each hour.  However, each observation of a field consisted of six scans.

[Referee]
> 5.e The authors may wish to specify in Section 4.3 that the SDSS DR9 Galactic Plane data
> that were used to test the IPHAS calibration were specifically from the SEGUE program
> of SDSS II, and cite the appropriate reference for SEGUE.

[Referee]
> 6.a. Section 5.1
> Would it be possible to provide the aperture corrections applied to the aperture
> photometry for each image along with the image metadata?

> Shouldn't the "saturated" column flag be specified per band?  Surely it must
> be possible for extremely red objects to be saturated in the i-band exposure
> but not r-band.

> 6.b. Section 5.2
> Little information about the specifics of band-merging is provided other than the
> reference to Taylor 2006 and TMATCHN.  Users will be forced to track down
> the reference to understand the process.  The authors should consider giving a short
> summary of the algorithm.  Some important information that should be provided are:

- How is confusion handled in the band merging process (e.g. what happens when
more than one i-band detection can be matched to an r-band detection?).
Source confusion in the band-merging process is mentioned in the final
paragraph of section 6.3.

- Is there any band-filling when a source is not detected in all bands?  If so, what
is the algorithm used to provide the flux limits in the non-detected bands?

6.c. Section 5.3 - Table 4

It is not clear what is the specific value is of the average magnitude given for each of
the different band-detection combinations in Table 4?  Wouldn't something like the peak
of the log N - log M curves be of more interest to provide some sense of the
completeness and/or depth of the Catalogue?

6.d. Section 5.4

While the authors provide a thorough description of the criteria used for
Catalogue selection, no reasons were offered for why those criteria were adopted.
For example, were there requirements for completeness and reliability that were
being satisfies?  If not, what were the objectives?

In the first paragraph of section 5.4, it is stated the three criteria were
enforced to limit the size of the catalogue.  Was this the main driver for establishing
the top-level criteria?  It seems unusual in this day and age of massive catalogues
that this would be a driver.

As stated earlier in this review, some description should be offered as to the nature of
the very large number of detections in the catalogue that are not classified as "reliable."
It would be useful to describe them to the user, and to offer suggestions of
how to recognize problematic detections, and also ones that may be real.

7. Section 6.1, par. 2, sentence 1

Should the work "accuracy" be "reliability"?

8. Figure 11 - caption

The reference to the left and right panels should presumably be to the top and bottom panels,
unless the two figures will be arranged differently in the published version.

### 2014-07-28 paper/referee/response-revision1.txt

[Referee]
> 1. In Section 3.1, paragraph 1, sentence 1:
> "archival" should be "archiving"

[Authors' response]
This has been corrected.

[Referee]
> 2. In the second paragraph of section 3.1, it is stated that a deblending routine
> is used to remove contaminating sources.  Additional information about this routine
> should be provided.  Are contaminants removed from the sky background measurement,
> the source measurement aperture, both?

[Authors' response]
The second paragraph of 3.1 has been edited to explain the steps of the CASU
object detection pipeline in more detail.

@@

[Referee]
> 3. At the beginning of the Data Reduction section (section 3), it would be helpful
> to provide a summary of how data processing for this release differed from that
> used for the Initial Data Release.

[Authors' response]
We have edited the final paragraph of Section 3.1 to highlight the key differences
between the present and previous data release.

@@

[Referee]
> 4.a. It is stated in section 3.2 that the exposures were discarded unless they
> had 5-sigma limiting magnitudes of r>20 or i>19 and H-apha>19 mag. How were the
> limiting magnitudes for an exposure estimated?

[Authors' response]
We have added a footnote to Section 3.2 which details how the 5-sigma limiting 
magnitude is defined and estimated in our work.

@@

[Referee]
> 4.b. Section 3.2 also specifies that an ellipticity limit of eps>0.3 was used.
> What is the justification of this limit?  For example, what is the potential impact on
> photometry and sensitivity at eps=0.25?

[Authors' response]
We have edited the third paragraph in Section 3.2 to explain that the choice
of our ellipticity limit at eps>0.3 is based on inspecting examples of 
high-ellipticity exposures, which revealed no evidence for degraded photometry
up to ellipticities less than 0.3. In fact high ellipticities are correlated 
with superior seeing, and therefore somewhat deeper frames.

### 2014-07-28 paper/iphas-dr2-rev1.tex
minor rev1 update
-

All raw IPHAS data were transferred
to the Cambridge Astronomical Survey Unit (CASU) 
for initial processing and archiving.
The procedures used by CASU were originally devised
for the INT Wide Field imaging Survey \citep[WFS;][]{McMahon2001,Irwin2005},
which was a 200 deg$^2$ survey programme carried out 
between 1998 and 2003.
Because IPHAS uses the same telescope and camera combination,
we have been able to benefit from the existing WFS pipeline.
A detailed description of the processing steps 
is found in \citet{Irwin2001}.
Its application to IPHAS has previously been described
by \citet{Drew2005} and \citet{Gonzalez-Solares2008},
and the source code is available
on-line\footnote{http://casu.ast.cam.ac.uk/surveys-projects/software-release}. 
In brief, the pipeline takes care of bias subtraction,
linearity correction, flat-fielding,
gain correction and de-fringing.

The reduced images are then stored in multi-extension FITS files
with a primary header describing the characteristics
(position, filter, exposure time, etc.) 
and four image extensions 
corresponding to each of the four CCDs.
Source detection and characterisation is then carried out 
using the \textsc{imcore} tool \citep{Irwin1985,Irwin1997}.
The flux of each source is measured using both
the peak pixel height
(i.e. a square 0\farcs 33$\times$0\farcs 33 aperture)
and a series of circular apertures of increasing diameter 
(1\farcs 2, 2\farcs 3, 3\farcs 3, 4\farcs 6 and 6\farcs 6).

The local background levels are estimated 
by computing the sigma-clipped median
flux in a grid of 64$\times$64 pixels (21\arcsec$\times$21\arcsec)
across the image,
which is then interpolated to obtain an estimate 
of the background level at each pixel.
These sky levels are subtracted from the aperture photometry and
-- when required --
a deblending routine is applied to remove
the contamination from any nearby sources.
This approach works very well 
across the vast majority of the survey area.  Nevertheless,
the Galactic Plane contains crowded regions 
with large numbers of overlapping sources
or rapidly spatially-varying nebulosity,
where aperture photometry will be compromised by frequent blending
or poor background subtraction.
In \S\ref{sec:catalogue} we will explain how affected sources are 
flagged.

@@

(1) \emph{Depth.} 
We discarded any exposures for which the $5\sigma$ limiting magnitude 
was brighter than 20th magnitude in the $r$-band
or brighter than 19th in $i$ or \ha. 
Such data were typically obtained during poor weather or full moon.
Most observations were significantly better than these limits.
Fig.~\ref{fig:depth} presents the distribution of limiting magnitudes
for all quality-approved fields;
the mean depths and standard deviations are 
$21.2\pm0.5$ ($r$), $20.0\pm0.3$ ($i$) and $20.3\pm0.3$ (\ha).
The depth achieved depended
most strongly on the presence of the moon,
which was above the horizon during 62 per cent 
of the observations.
The great range in sky brightness this
produced is behind the wide and bi-modal shape
of the $r$-band limiting magnitude distribution 
(top panel in Fig.~\ref{fig:depth}).
In contrast, the depths attained in $i$ and \ha\ 
are less sensitive to moonlight, leading to
narrower magnitude limit distributions
(middle and bottom panels in Fig.~\ref{fig:depth}).
To a lesser extent, the wide spread in the $r$-band depth
is also explained by the shorter exposure time
that was used for this band during the first months of data-taking.

(2) \emph{Ellipticity.} 
The ellipticity of a point source,
defined as $e = 1 - b / a$ 
with $b$ the semi-minor and $a$ the semi-major axis,
is a morphological measure of the elongation
of the Point Spread Function (PSF).
It is expected to be zero (circular)
in a perfect imaging system,
but it is slightly non-zero in any real telescope data 
due to optical distortions and tracking errors.
Indeed, it is worth noting
that IPHAS data have been collected
from unguided exposures that rely entirely
on the INT's tracking capability.
The mean ellipticity measured
in the data is $0.09\pm0.04$.
There have been sporadic episodes with higher ellipticities
due to mechanical glitches in the telescope tracking system.
To exclude these, we rejected exposures
in which the \emph{average} ellipticity
across the detectors exceeded $e > 0.3$.
We have also found that, as this threshold is breached,
the photometric measurements delivered by the pipeline become degraded.

>>>>

All raw IPHAS data were transferred
to the Cambridge Astronomical Survey Unit (CASU) 
for initial processing and archiving.
The procedures used by CASU were originally devised
for the INT Wide Field imaging Survey \citep[WFS;][]{McMahon2001,Irwin2005},
which was a 200 deg$^2$ survey programme carried out 
between 1998 and 2003.
Because IPHAS uses the same telescope and camera combination,
we have been able to benefit from the existing WFS pipeline.
A description of the processing steps 
is found in \citet{Irwin2001}.
Its application to IPHAS has previously been described
by \citet{Drew2005} and \citet{Gonzalez-Solares2008},
and the source code is available
on-line\footnote{http://casu.ast.cam.ac.uk/surveys-projects/software-release}. 
In brief, the pipeline takes care of bias subtraction,
linearity correction, flat-fielding,
gain correction and de-fringing.

%The reduced images are then stored in multi-extension FITS files
%with a primary header describing the characteristics
%(position, filter, exposure time, etc.) 
%and four image extensions 
%corresponding to each of the four CCDs.

% called \textsc{imcore},

{ \bf
Source detection and parametrisation is then carried out
using the standard methods in use by CASU,
which can be summarised in four steps
\citep[a discussion on each of these steps
and related points can be found in][]{Irwin1985,Irwin1997}:
\begin{enumerate}
\item The local sky background is estimated
by computing the sigma-clipped median intensity
in a grid of 64$\times$64 pixel bins across the image,
which is robust against contaminating sources.
Bilinear interpolation is then used to obtain an estimate 
of the background level at each pixel.
\item Objects, or blends of objects, are detected
by identifying groups of 4 or more neighbouring pixels
in which the intensity exceeds the background level by $1.25\sigma$.
\item The object are parametrised:
positions are obtained by computing 
the intensity-weighted isophotal centre-of-gravity
of each object,
while photometry is obtained 
by measuring the intensity
in a series of soft-edged circular apertures of increasing diameter 
(1\farcs 2, 2\farcs 3, 3\farcs 3, 4\farcs 6 and 6\farcs 6).
\item Finally, the parametrisation of overlapping objects
is refined by simultaneous fitting of
soft-edged apertures to each blend,
which is essentially identical 
to multiple "profile" fitting.
\end{enumerate}
A discussion on each of these steps
can be found in \citep{Irwin1985,Irwin1997}.
We note that the parametrisation of single, unconfused objects
is naturally more reliable than that of
objects which had to be deblended.
In \S\ref{sec:catalogue} we will explain
how blended sources are flagged in the catalogue.

%The sentence about deblending could be misleading since the objects are
%detected using our standard deblending algorithm while the aperture flux
%estimates for overlapping images are computed by simultaneous fitting of
%soft-edged apertures.  During this multiple "profile" fitting the sky
%level is based on the interpolated global detector-level fit

%This approach works very well 
%across the vast majority of the survey area.  Nevertheless,
%the Galactic Plane contains crowded regions 
%with large numbers of overlapping sources
%or rapidly spatially-varying nebulosity,
%where aperture photometry will be compromised by frequent blending
%or poor background subtraction. 
}

@@

(1) \emph{Depth.} 
We discarded any exposures 
for which the $5\sigma$ limiting magnitude\footnote{\bf
We defined the $5\sigma$ limiting magnitude
as the magnitude which a point source would have,
when its flux in a 2\farcs 3-diameter aperture
is five times the level of the noise in the sky background.
The sky noise was estimated using a robust MAD estimator 
for noise scaled to equivalent Gaussian rms value, ie. = MAD x 1.48,
after removing large scale sky background variations.
(MAD = Median of the Absolute Deviations about the median.)}
was brighter than 20th magnitude in the $r$-band
or brighter than 19th in $i$ or \ha.
Such data were typically obtained during poor weather or full moon.
Most observations were significantly better than these limits.
Fig.~\ref{fig:depth} presents the distribution of limiting magnitudes
for all quality-approved fields;
the mean depths and standard deviations are 
$21.2\pm0.5$ ($r$), $20.0\pm0.3$ ($i$) and $20.3\pm0.3$ (\ha).
The depth achieved depended
most strongly on the presence of the moon,
which was above the horizon during 62 per cent 
of the observations.
The great range in sky brightness this
produced is behind the wide and bi-modal shape
of the $r$-band limiting magnitude distribution 
(top panel in Fig.~\ref{fig:depth}).
In contrast, the depths attained in $i$ and \ha\ 
are less sensitive to moonlight, leading to
narrower magnitude limit distributions
(middle and bottom panels in Fig.~\ref{fig:depth}).
To a lesser extent, the wide spread in the $r$-band depth
is also explained by the shorter exposure time
that was used for this band during the first months of data-taking.

(2) \emph{Ellipticity.} 
The ellipticity of a point source,
defined as $e = 1 - b / a$ 
with $b$ the semi-minor and $a$ the semi-major axis,
is a morphological measure of the elongation
of the Point Spread Function (PSF).
It is expected to be zero (circular)
in a perfect imaging system,
but it is slightly non-zero in any real telescope data 
due to optical distortions and tracking errors.
Indeed, it is worth noting
that IPHAS data have been collected
from unguided exposures that rely entirely
on the INT's tracking capability.
The mean ellipticity measured
in the data is $0.09\pm0.04$.
There have been sporadic episodes with higher ellipticities
due to mechanical glitches in the telescope tracking system.
{ \bf As a result, 3 per cent of our images show an
average ellipticity across the detectors which is worse
than $e > 0.2$.
The inspection of these examples revealed no evidence for
degraded photometry up to ellipticities less than 0.3.
In fact high ellipticities are correlated with superior seeing,
and therefore somewhat deeper frames.
We did however exclude a small number of exposures
which exceeded $e > 0.3$.}

%To exclude these, we rejected exposures
%in which the average ellipticity
%across the detectors exceeded $e > 0.3$.
%We have also found that, as this threshold is breached,
%the photometric measurements delivered by the pipeline become degraded.
